### 2021-10-27 article.tex
Shaping up the introduction
s/cite/citep/g

We believe that type-level programming is not necessarily incompatible with other programming paradigms and that the current division exists mainly due to a lack of attention from the research community.
Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\cite{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.

@@

In this section, we formally present \SystemFm, an extension of \SystemFsub~\cite{cardelli1994extension} with pattern matching, opaque classes, and match types.
\Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
We discuss differences to \SystemFsub in the following subsections (\Cref{subsec:classes} and \ref{subsec:matches}).
In \Cref{subsec:type-safety}, we outline a proof of type safety for \SystemFm.
In \Cref{subsec:type-binding-extension}, we present an extension of \SystemFm with support for binding pattern variables in type patterns.

@@

\SystemFm supports both pattern matching on the term level (\emph{match expressions}) as well as on the type-level (\emph{match types}).
Matches, both on terms and on types, are composed of a scrutinee, a list of cases and a default expression/type.
Each case consists of a \emph{type test} and a corresponding expression/type.
At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \cite{abadi1991dynamic}).
At the type level, a type test corresponds to a subtyping test with a particular type.
This disparity reflects the difference between runtime, where type tests are implemented using class tables, and compile time, where types are compared using the type system in its full extent.
We discuss the representation of Scala types at runtime in \Cref{subsec:types-at-runtime}.

@@

\Cref{fig:structure} gives an overview of the proof structure by showing implications between the various lemmas and theorems.
The basic structure resembles that of \SystemFsub's standard safety proof from~\cite{pierce2002types}.
We continue our presentation by introducing the lemmas and theorems used in our type safety proof.

@@

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \cite{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.

@@

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \cite{fengyun2016generic}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

@@

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \cite{schinz2005compiling}.
For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \cite{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.

@@

\SystemFm's type rules enable any match expression to be typed as a match type, but the situation is different in the full Scala language.
Pattern matching in Scala supports many sorts of patterns~\cite{emir2007matching}, most of which do not have a match type counterpart.
Furthermore, typing match expressions as match types is not enabled by default in order to preserve backward compatibility.
Instead, explicit type annotations must be provided.

@@

In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\cite{harris2020numpy}.
The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
Shape and indexing errors in ndarrays is a widely acknowledged problem~\cite{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\cite{chen2017typesafe, huang2017hasktorch}.
Our library uses match types to provide a shape-safe NumPy-like interface.
We believe that the approach presented in this case study is an improvement over the status quo because it does not use metaprogramming or any sort of convoluted encoding to express type-level operations.

% \subsection{Related Work}
% Shape and indexing errors in ndarrays is a widely acknowledged problem~\cite{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
% The Dex language~\cite{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
% Nexus~\cite{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
% However, it only guards against rank errors, and not ndarray size errors.

% Match types offer an abstraction that can be used to express what these solutions cannot. They allow us to program shape transformations and type constraints, in order to  achieve shape-safety for variable ranks.
% The implementation in this case study is inspired by the HaskTorch Haskell library~\cite{huang2017hasktorch}, which also represents shapes using HLists, and makes direct use of type-level computation to compute shape transformations.

@@

Scala supports singleton types, which are types inhabited by a single value~\cite{leontiev2014sip}.
For instance, the singleton type \dotty{1} denotes the type containing the integer value \dotty{1}.
The Scala standard library contains several predefined match types to perform arithmetic operations at the type level.
For instance, \dotty{type +[A <: Int,}\dotty{B <: Int]} represents the addition of integer singleton types.
Internally, the compiler is special-cased to implement these arithmetic operations using constant folding.
Representing numbers with singleton types is desirable for practical purposes, but not absolutely necessary for this case study (for instance, Peano numerals could be used instead).

@@

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\cite{kiselyov2004strongly}.
We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
%
\shapecodesection
%
This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.

To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
%
\ndarraycodesection

@@

To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalcodesection

@@

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplycodesection
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

@@

An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementscodesection

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
%
\reshapecodesection
%
With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

@@

\npmeancodesection

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reducecodesection

@@

Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
%
\badreducescalacodesection
%
As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.

% \goodreducescalacodesection

@@

Dependently typed calculi typically use the same language to describe terms and types.
This unification is also commonly used in the presence of subtyping~\cite{zwanenburg1999pure, hutchins2010pure, yang2017unifying}.
For systems with a complete term/type symmetry, this is a natural design, as it is concise and simplifies the meta-theory.
Unfortunately, the lack of distinction between term and type level renders these systems impractical for our purpose, given that our research takes place in the context of an existing language with a clear term/type separation.

Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\cite{aspinall1994singleton, stone2000deciding, courant2003strong}.
Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
This mechanism is appealing because it allows type system designers to cherry-pick the term constructs that should be allowed in types.
When multiple constructs are shared between terms and types, singleton types provide a clear economy of concepts.
In our study of match types, a minimal use of singleton types would result in sharing a single constructor between the term and the type languages: the constructor for matches. It is unclear if the benefits in doing so would outweigh the additional complexity.

Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\cite{essence2016amin}.
DOT does not directly support any form of type-level computation.
We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\cite{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.

@@

In their work on intensional type analysis~\cite{harper1995compiling}, Harper and Morrisett introduce the $\λ^{ML}_i$ calculus that supports structural analysis of types.
In $\λ^{ML}_i$, types are represented as expressions that can be inspected by case analysis using a "typecase" construct, available both at the term and at the type level.
Match types can be seen as an extension of intensional type analysis to work with object-oriented class hierarchies and subtyping.
Whereas patterns in $\λ^{ML}_i$ are limited to a fixed set of disjoint types, match types need to deal with open class hierarchies, of which not all members are known at compile-time.
This means pattern types can overlap, and we need to perform an analysis of disjointness between the scrutinee type and each pattern type.
Disjointness allows for sound reduction in the presence of overlapping patterns and abstract scrutinee types, while retaining the natural sequential evaluation order of pattern matching.

@@

Haskell's type families allow programmers to define type-level functions using pattern matching~\cite{chakravarty2005associated, schrijvers2008open}.
By default, type families are open, which means that a definition can spread across multiple files and compilations.
This flexibility induces a substantial restriction on type family definitions: patterns must not overlap.
One benefit of this restriction is that it prevents any ambiguity in the reduction of type families (patterns are pairwise disjoint), which is required given the distributed nature of definitions.
Open type families are well-suited to be used in conjunction with type classes, since both constructs have open-ended definitions with non-overlapping constraints.

@@

By default, Haskell checks for termination of recursive type families, but this check can be disabled to increase type families' expressiveness.
Although the formalization presented in~\cite{eisenberg2014closed} does not cover non-terminating families, the paper discusses a soundness problem caused by non-termination.
The problem only occurs in the presence of repeated type bindings in patterns.
In Scala, match types (and pattern matching in general) do not allow repeated bindings and are therefore not affected by this problem.

% A significant difference between type-checking type families and match types is that type families are entirely handled during type inference, while match types reduce during subtyping.
% Influencing Haskell's global type inference is notoriously difficult, which can, unfortunately, hinder the usability of type families.
% For instance, in a functional pearl on typed matrices~\cite{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.

@@

In their introductory example,~\citeauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:
%
\begin{lstlisting}
trait AgeClass {
  type Age
  def addAge(a: Age, i: Int): Int
}
object AgeObject extends AgeClass {
  type Age = Int
  def addAge(a: Age, i: Int): Int = a + i
}
\end{lstlisting}
%
In this example, type \dotty{Age} is abstract in \dotty{AgeClass} and concrete in \dotty{AgeObject}.
In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates \dotty{Age} and \dotty{Int}.
Such type family would reduce differently in \dotty{AgeClass}, where the types are different, and in \dotty{AgeObject}, where those two types are synonyms, which can easily be exploited to obtain a runtime error.

@@

The TypeScript language specification briefly describes the algorithm used to reduce conditional types in the presence of type variables~\cite{microsoft2020typescript}.
Given a type \typescript{S extends T ? Tt : Tf}, the TypeScript compiler first replaces all the type parameters in \typescript{S} and \typescript{T} by \typescript{any} (the top of TypeScript's subtyping lattice).
If the resulting types (after substitution) are not subtypes, the overall condition is reduced to \typescript{Tf}.
Unfortunately, this algorithm is both \emph{unsound} and \emph{incomplete}.

@@

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\cite{bierman2014typescript}.
Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.

>>>>

We believe that type-level programming is not necessarily incompatible with other programming paradigms and that the current division exists mainly due to a lack of attention from the research community.
Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\citep{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.

@@

In this section, we formally present \SystemFm, an extension of \SystemFsub~\citep{cardelli1994extension} with pattern matching, opaque classes, and match types.
\Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
We discuss differences to \SystemFsub in the following subsections (\Cref{subsec:classes} and \ref{subsec:matches}).
In \Cref{subsec:type-safety}, we outline a proof of type safety for \SystemFm.
In \Cref{subsec:type-binding-extension}, we present an extension of \SystemFm with support for binding pattern variables in type patterns.

@@

\SystemFm supports both pattern matching on the term level (\emph{match expressions}) as well as on the type-level (\emph{match types}).
Matches, both on terms and on types, are composed of a scrutinee, a list of cases and a default expression/type.
Each case consists of a \emph{type test} and a corresponding expression/type.
At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \citep{abadi1991dynamic}).
At the type level, a type test corresponds to a subtyping test with a particular type.
This disparity reflects the difference between runtime, where type tests are implemented using class tables, and compile time, where types are compared using the type system in its full extent.
We discuss the representation of Scala types at runtime in \Cref{subsec:types-at-runtime}.

@@

\Cref{fig:structure} gives an overview of the proof structure by showing implications between the various lemmas and theorems.
The basic structure resembles that of \SystemFsub's standard safety proof from~\citep{pierce2002types}.
We continue our presentation by introducing the lemmas and theorems used in our type safety proof.

@@

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \citep{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.

@@

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \citep{fengyun2016generic}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

@@

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \citep{schinz2005compiling}.
For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.

@@

\SystemFm's type rules enable any match expression to be typed as a match type, but the situation is different in the full Scala language.
Pattern matching in Scala supports many sorts of patterns~\citep{emir2007matching}, most of which do not have a match type counterpart.
Furthermore, typing match expressions as match types is not enabled by default in order to preserve backward compatibility.
Instead, explicit type annotations must be provided.

@@

In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\citep{harris2020numpy}.
The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\citep{chen2017typesafe, huang2017hasktorch}.
Our library uses match types to provide a shape-safe NumPy-like interface.
We believe that the approach presented in this case study is an improvement over the status quo because it does not use metaprogramming or any sort of convoluted encoding to express type-level operations.

% \subsection{Related Work}
% Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
% The Dex language~\citep{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
% Nexus~\citep{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
% However, it only guards against rank errors, and not ndarray size errors.

% Match types offer an abstraction that can be used to express what these solutions cannot. They allow us to program shape transformations and type constraints, in order to  achieve shape-safety for variable ranks.
% The implementation in this case study is inspired by the HaskTorch Haskell library~\citep{huang2017hasktorch}, which also represents shapes using HLists, and makes direct use of type-level computation to compute shape transformations.

@@

Scala supports singleton types, which are types inhabited by a single value~\citep{leontiev2014sip}.
For instance, the singleton type \dotty{1} denotes the type containing the integer value \dotty{1}.
The Scala standard library contains several predefined match types to perform arithmetic operations at the type level.
For instance, \dotty{type +[A <: Int,}\dotty{B <: Int]} represents the addition of integer singleton types.
Internally, the compiler is special-cased to implement these arithmetic operations using constant folding.
Representing numbers with singleton types is desirable for practical purposes, but not absolutely necessary for this case study (for instance, Peano numerals could be used instead).

@@

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
%
\shapeCodeSection
%
This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.

To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
%
\ndarrayCodeSection

@@

To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalCodeSection

@@

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplyCodeSection
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

@@

An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementsCodeSection

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
%
\reshapeCodeSection
%
With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

@@

\npmeanCodeSection

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reduceCodeSection

@@

Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
%
\badreducescalaCodeSection
%
As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.

% \goodreducescalaCodeSection

@@

Dependently typed calculi typically use the same language to describe terms and types.
This unification is also commonly used in the presence of subtyping~\citep{zwanenburg1999pure, hutchins2010pure, yang2017unifying}.
For systems with a complete term/type symmetry, this is a natural design, as it is concise and simplifies the meta-theory.
Unfortunately, the lack of distinction between term and type level renders these systems impractical for our purpose, given that our research takes place in the context of an existing language with a clear term/type separation.

Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\citep{aspinall1994singleton, stone2000deciding, courant2003strong}.
Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
This mechanism is appealing because it allows type system designers to cherry-pick the term constructs that should be allowed in types.
When multiple constructs are shared between terms and types, singleton types provide a clear economy of concepts.
In our study of match types, a minimal use of singleton types would result in sharing a single constructor between the term and the type languages: the constructor for matches. It is unclear if the benefits in doing so would outweigh the additional complexity.

Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\citep{essence2016amin}.
DOT does not directly support any form of type-level computation.
We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\citep{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.

@@

In their work on intensional type analysis~\citep{harper1995compiling}, Harper and Morrisett introduce the $\λ^{ML}_i$ calculus that supports structural analysis of types.
In $\λ^{ML}_i$, types are represented as expressions that can be inspected by case analysis using a "typecase" construct, available both at the term and at the type level.
Match types can be seen as an extension of intensional type analysis to work with object-oriented class hierarchies and subtyping.
Whereas patterns in $\λ^{ML}_i$ are limited to a fixed set of disjoint types, match types need to deal with open class hierarchies, of which not all members are known at compile-time.
This means pattern types can overlap, and we need to perform an analysis of disjointness between the scrutinee type and each pattern type.
Disjointness allows for sound reduction in the presence of overlapping patterns and abstract scrutinee types, while retaining the natural sequential evaluation order of pattern matching.

@@

Haskell's type families allow programmers to define type-level functions using pattern matching~\citep{chakravarty2005associated, schrijvers2008open}.
By default, type families are open, which means that a definition can spread across multiple files and compilations.
This flexibility induces a substantial restriction on type family definitions: patterns must not overlap.
One benefit of this restriction is that it prevents any ambiguity in the reduction of type families (patterns are pairwise disjoint), which is required given the distributed nature of definitions.
Open type families are well-suited to be used in conjunction with type classes, since both constructs have open-ended definitions with non-overlapping constraints.

@@

By default, Haskell checks for termination of recursive type families, but this check can be disabled to increase type families' expressiveness.
Although the formalization presented in~\citep{eisenberg2014closed} does not cover non-terminating families, the paper discusses a soundness problem caused by non-termination.
The problem only occurs in the presence of repeated type bindings in patterns.
In Scala, match types (and pattern matching in general) do not allow repeated bindings and are therefore not affected by this problem.

% A significant difference between type-checking type families and match types is that type families are entirely handled during type inference, while match types reduce during subtyping.
% Influencing Haskell's global type inference is notoriously difficult, which can, unfortunately, hinder the usability of type families.
% For instance, in a functional pearl on typed matrices~\citep{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.

@@

In their introductory example,~\citepauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:
%
\begin{lstlisting}
trait AgeClass {
  type Age
  def addAge(a: Age, i: Int): Int
}
object AgeObject extends AgeClass {
  type Age = Int
  def addAge(a: Age, i: Int): Int = a + i
}
\end{lstlisting}
%
In this example, type \dotty{Age} is abstract in \dotty{AgeClass} and concrete in \dotty{AgeObject}.
In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates \dotty{Age} and \dotty{Int}.
Such type family would reduce differently in \dotty{AgeClass}, where the types are different, and in \dotty{AgeObject}, where those two types are synonyms, which can easily be exploited to obtain a runtime error.

@@

The TypeScript language specification briefly describes the algorithm used to reduce conditional types in the presence of type variables~\citep{microsoft2020typescript}.
Given a type \typescript{S extends T ? Tt : Tf}, the TypeScript compiler first replaces all the type parameters in \typescript{S} and \typescript{T} by \typescript{any} (the top of TypeScript's subtyping lattice).
If the resulting types (after substitution) are not subtypes, the overall condition is reduced to \typescript{Tf}.
Unfortunately, this algorithm is both \emph{unsound} and \emph{incomplete}.

@@

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014typescript}.
Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.
